---
title: "Car thefts rates have tripled in Baltimore. We visited the hardest-hit neighborhood."
author: "Greg Morton"
date: "2023-11-10"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
# loading packages 
library(tidyverse)
library(tidycensus)
library(lubridate)
library(janitor)
library(sf)
library(mapview)
library(kableExtra)

# Turn off scientific notation 
options(scipen=999)

# Most recent part 1 crimes data 
# download here: https://data.baltimorecity.gov/search?q=crime%20data
# use init.qmd in data/car-thefts to reformat P1 crimes create "victims" object 
victims <- read_csv("data/victims.csv")

# Baltimore neighborhoods shapefile 
neighbs <- st_read("data/neighborhoods/Neighborhoods.shp")

# Car thefts (2023)
auto_theft_2023 <- 
  victims %>% 
  filter(str_detect(str_to_upper(description), "AUTO THEFT")) %>% 
  filter(crime_date_time >= "2023-01-01")


# Car thefts (2021 - 2023)
auto_theft_yoy <- 
  victims %>% 
  filter(str_detect(str_to_upper(description), "AUTO THEFT")) %>% 
  filter(crime_date_time >= "2021-01-01") %>% 
  filter(!is.na(x) | !is.na(y))

### Section 1: Examining car thefts by neighborhood

# turn p1 crime data into a simple features object
auto_theft_2023_sf <- 
  auto_theft_2023 %>% 
  filter(!is.na(x) | !is.na(y)) %>% 
  st_as_sf(coords = c("x", "y"),
           crs = 4326) %>% 
  mutate(location = 
           case_when(str_detect(location, " rd$") ~ str_replace(location, " rd$", " road"),
                     str_detect(location, " avenue$") ~ str_replace(location, " avenue$", " ave"),
                     str_detect(location, " street$") ~ str_replace(location, " street$", " st"),
                     str_detect(location, " cir$") ~ str_replace(location, " cir$", " circle"),
                     str_detect(location, " cir n$") ~ str_replace(location, " cir n$", " circle n"),
                     str_detect(location, " cir s$") ~ str_replace(location, " cir s$", " circle s"),
                              TRUE ~ location))



auto_theft_since_2021_sf <-
  st_as_sf(auto_theft_yoy,
           coords = c("x", "y"),
           crs = 4326) %>%
  mutate(location = 
           case_when(str_detect(location, " rd$") ~ str_replace(location, " rd$", " road"),
                     str_detect(location, " avenue$") ~ str_replace(location, " avenue$", " ave"),
                     str_detect(location, " street$") ~ str_replace(location, " street$", " st"),
                     str_detect(location, " cir$") ~ str_replace(location, " cir$", " circle"),
                     str_detect(location, " cir n$") ~ str_replace(location, " cir n$", " circle n"),
                     str_detect(location, " cir s$") ~ str_replace(location, " cir s$", " circle s"),
                              TRUE ~ location)) 

# intersect crime data and neighborhood data 
crime_by_neighb_geo <- st_transform(auto_theft_2023_sf, st_crs(neighbs))
crime_by_neighb_geo_since_2021 <- st_transform(auto_theft_since_2021_sf, st_crs(neighbs)) 


# Perform a spatial join to find which points in fall within which neighborhood boundaries
thefts_in_neighbs <- st_join(crime_by_neighb_geo, neighbs)
thefts_in_neighbs_since_2021 <- st_join(crime_by_neighb_geo_since_2021, neighbs)

# YOY change by neighborhood
yoy_change_neighb <-
  thefts_in_neighbs_since_2021 %>% 
  # filter out neighborhood mismatches where geolocation neighborhood doesn't match listed neighborhood
  filter(str_to_upper(name) == str_to_upper(neighborhood)) %>% 
  group_by(name,
           crime_date = case_when(crime_date_time >= "2021-01-01" & crime_date_time <= "2021-11-07" ~ "2021",
                                  crime_date_time >= "2022-01-01" & crime_date_time <= "2022-11-07" ~ "2022",
                                  crime_date_time >= "2023-01-01" ~ "2023",
                                  TRUE ~ "no year"
                                  )) %>% count() %>% 
  pivot_wider(names_from = "crime_date", values_from = "n") %>% 
  ungroup() %>% 
  group_by(name) %>% 
  summarise(x2021 = sum(`2021`, na.rm = T ),
            x2022 = sum(`2022`, na.rm = T),
            x2023 = sum(`2023`, na.rm = T),
            difference_raw = x2023 - x2022) %>% 
  arrange(desc(x2023)) 

### Section 2: Examining Car thefts by block 

# block intersections (for all auto thefts since 2022)
intersections <-
  read_csv("data/block_intersections.csv")


# check for block intersections on all auto thefts 
auto_theft_since_2022_corrected <- 
  thefts_in_neighbs_since_2021 %>% 
  filter(crime_date_time >= "2022-01-01") %>% 
  filter(!is.na(geometry)) %>% 
  filter(str_to_upper(name) == str_to_upper(neighborhood)) %>%
  left_join(intersections) %>% 
  filter(intersects == TRUE) %>% 
  distinct(cc_number, gender, age, .keep_all = TRUE) 


```

"As of Nov. 7, 9,523 auto theft incidents --- car thefts or attempted auto thefts that can render cars heavily damaged and inoperable --- were reported in the city, which is more than triple the number of cars stolen from the same time period last year."

```{r}
# NOTE: Data is as of Nov. 7 2023

# car thefts (2023)
auto_theft_2023 %>% 
  group_by(year = floor_date(crime_date_time, "year")) %>% 
  count() %>% 
  kbl() %>% 
  kable_material_dark()
```

"More than 1,000 cars have been stolen in Baltimore every month since June."

```{r pressure, echo=FALSE}
# NOTE: Data is as of Nov. 7 2023

auto_theft_mom <-
  victims %>% 
  filter(str_detect(str_to_upper(description), "AUTO THEFT")) %>% 
  filter(crime_date_time >= "2023-01-01") %>% 
  group_by(floor_date(crime_date_time, "month")) %>% 
  count() %>% 
  kbl() %>% 
    kable_material_dark() %>% 
    row_spec(6:10, bold = T, color = "white", background = "#1f937e")

auto_theft_mom 
```

"During the peak of the auto theft crisis in July, police responded to over 25 auto theft incidents per 10,000 residents, more, by far, than any other county in the state."

```{r}


md_per_10k <-
  read_csv("data/md_per_capita.csv")

md_per_10k %>% 
  filter(month > "2022-12-01" & month <= "2023-10-01") %>% 
  kbl() %>% 
  kable_material_dark() %>% 
  column_spec(2, bold = T, color = "black", background = "#ffffdd")
  
```

"In Frankford alone, there have been 350 reported auto theft incidents so far this year, more than in any other city neighborhood and an increase of 332% from the same time last year."

"Nearby Belair-Edison had the second-most-stolen auto incidents, with more than 300, up from 78 this time last year."

```{r}

neighborhood_change <- 
  yoy_change_neighb %>% 
    as.data.frame() %>% 
    select(-geometry) %>% 
    head(25) 


neighborhood_change %>% 
    kbl() %>% 
    kable_material_dark() %>% 
    row_spec(1, bold = T, color = "white", background = "#4d2d8e") %>% 
    row_spec(2, bold = T, color = "white", background = "#795eab")



```

"Those blocks, which include the 4900-5100 blocks of Goodnow Road and 4900 block of Gunther Avenue, saw seven reported auto thefts incidents all of last year. This year, there have already been 43."

```{r}
# Goodnow/Gunther auto theft incidents since 2022
goodnow_rd <- auto_theft_since_2022_corrected %>%  
  filter(location == "5000 goodnow road" | location == "4900 goodnow road" | location == "5100 goodnow road" | location == "4900 gunther ave") %>% 
  select(crime_date_time, time_range, crime_code, location, neighborhood)

# YOY change
goodnow_chart <- 
  goodnow_rd %>% 
    as.data.frame() %>% 
    group_by(year = floor_date(crime_date_time, "year")) %>% 
    count() %>% 
    kbl() 

goodnow_chart %>% 
    kable_material_dark() %>% 
    row_spec(2, bold = T, color = "white", background = "#a3c4bd")


# Map Goodnow Road/Gunther Ave auto theft incidents 
goodnow_map <- 
  mapview(goodnow_rd)

goodnow_map


```

"From July 23 to 25, and again from Aug. 26 to 29, an auto theft incident was reported on those blocks every day."

```{r}

goodnow_daily_auto_theft <- goodnow_rd %>% 
  filter(crime_date_time > "2022-12-31") %>% 
  arrange(crime_date_time) %>% 
  as.data.frame() %>% 
  select(-geometry)

goodnow_daily_auto_theft %>%   
  kbl() %>% 
  kable_material_dark() %>% 
  row_spec(c(27:29, 24:25, 39:42, 2:3), bold = T, color = "white", background = "#a75bb3") %>% 
  row_spec(c(16:19, 30:36), bold = T, color = "white", background = "#d1bcf4")
  



```

"According to census data, about two-thirds of residents drive to work, about 20% higher than the average for the rest of the city. Residents commute around 30 minutes to their jobs on average"

```{r}

commute_data <- get_acs(
  geography = "tract",
  state = "MD",
  county = "Baltimore City",
  year = 2021,
  variables = c(pop = "B01001_001", 
                commuters = "B08006_001",
                drove_alone = "B08006_003",
                public_transit = "B08006_008",
                biked = "B08006_014",
                walked = "B08006_015",
                other = "B08006_016",
                worked_from_home = "B08006_017",
                carpool = "B08006_004",
                aggregate_commute_time = "B08013_001"
                ),
  geometry = TRUE
  ) %>% 
  st_transform(st_crs(goodnow_rd)) %>% 
  pivot_wider(names_from = "variable", 
              values_from = "estimate") %>% 
  group_by(GEOID)  %>% 
  summarise(pop = sum(pop, na.rm = T),
            commuters = sum(commuters, na.rm =T),
            drove_alone = sum(drove_alone, na.rm =T),
            carpool = sum(carpool, na.rm =T),
            public_transit = sum(public_transit, na.rm =T),
            biked = sum(biked, na.rm =T),
            walked = sum(walked, na.rm =T),
            other = sum(other, na.rm =T),
            worked_from_home = sum(worked_from_home, na.rm =T),
            aggregate_commute_time = sum(aggregate_commute_time, na.rm =T)
            ) %>% 
  mutate(pct_drive_alone = drove_alone / commuters * 100,
         sum = drove_alone + public_transit + biked + walked + other + worked_from_home + carpool,
         avg_commute_time = aggregate_commute_time / (commuters - worked_from_home),
         completeness = sum / commuters)


goodnow_demos <-
  st_join(goodnow_rd, commute_data) %>% 
  distinct(GEOID, .keep_all = TRUE) %>% 
  select(GEOID, 
         neighborhood,
         pct_drive_alone, 
         avg_commute_time,
         commuters, 
         drove_alone, 
         carpool, 
         public_transit, 
         biked,
         walked, 
         other, 
         worked_from_home, 
         completeness, 
         aggregate_commute_time) 

goodnow_demos %>% 
  kbl() %>% 
  kable_material_dark() %>% 
  column_spec(3:4, bold = T, color = "black", background = "#b8d53f")
  


```
